Fast-Match: Fast and robust feature matching on large images

### 2014-06-18 paper.tex
Hopefully last batch of corrections

In this paper we introduce a matching algorithm designed to match features between two images only in image areas that are likely to correspond. This approach is much faster than traditional methods because there is no need to compute descriptors for areas in the image that are not matched. We provide two variations of the algorithm: The \emph{general} variation functions as a traditional matching algorithm and matches two unknown images albeit faster and more robustly than conventional methods. The \emph{retrieval} variation on the other hand assumes that we know one of the images we intend to match beforehand; under this assumption, it can be a magnitude faster than existing matching methods without compromising on accuracy. We will unimaginatively refer to the proposed algorithm as \emph{Fast-Match} in this paper and make clear from the context whether this refers to the \emph{retrieval} or \emph{general} variation. As an aside we adopt an artificial distinction between the words \emph{match} and \emph{correspondence} to make it clear when a match between two images is correct. The word \emph{match} will be used to denote a result from a matching algorithm that may or may not be correct while a \emph{correspondence} refers to an actual visual correspondence between two points in two images.

@@

As noted, the cost of computing SIFT keypoints and descriptors is addressed by other local image features such as SURF \cite{bay2006surf}, BRIEF \cite{calonder2010brief}, or BRISK \cite{leutenegger2011brisk}, just to mention a few. Similarly efforts have been made to improve SIFT itself such as PCA-SIFT \cite{ke2004pca} and GLOH \cite{mikolajczyk2005performance}. Both apply PCA to reduce descriptor lengths and improve distinctiveness, but neither have been shown to consistently outperform SIFT 
\cite{mikolajczyk2005performance,li2008comprehensive}.

@@

A wealth of matching methods have focused instead on increasing the efficiency of local feature matching. \emph{Fast-Match} builds upon the foundation of \emph{Ratio-Match} introduced originally by Deriche et al.~\cite{deriche1994robust} and Baumberg~\cite{baumberg2000reliable} even though Lowe is usually credited for introducing it~\cite{lowe2004sift}. They both propose using the ratio of the similarity of the best to second best correspondence of a given point to evaluate how unique it is. Their finding has later been tested by several independent teams, all concluding that thresholding based on this ratio is generally superior to thresholding based on similarity or returning all nearest neighbours \cite{lowe2004sift,mikolajczyk2005performance,moreels2007evaluation,rabin2009statistical}.

Brown and Lowe \cite{brown2005multi} extend ratio match to deal with a set of images by using not the ratio of the best and second best correspondence, but the average ratio of the best and the average of second best correspondences across a set of images.  Rabin et al.\ \cite{rabin2009statistical} try to enhance descriptor matching by looking at the statistical distribution of local features in the matched images, and only return a match when such a correspondence would not occur by mere chance. Finally, Arnfred et al.\ generalize \emph{Ratio-Match} to make use of both of the matched images to provide a more accurate evaluation of the uniqueness of a match \cite{arnfred2013mirror}.

@@

For sparse local image features many solutions have combined \emph{Ratio-Match} with various geometric constraints to improve matching. These constraints are based on assumptions regarding the transformation between the \emph{query} and \emph{target images}. A commonly used example is \emph{RANSAC} applied to feature matching where matches are chosen from a pool of candidates according to how well they approximate a global epipolar geometry \cite{fischler1981ransac,torr2000mlesac,szeliski2010}. Similar global angular and distance constraints can be used to filter a set of matches as shown in \cite{kim2008efficient,schmid1997local}. Finally the problem of feature matching can be modelled as a graph matching problem where each feature is a vertex, and edge values correspond to a geometric relation between two features as demonstrated in \cite{torresani2008feature,yarkony2010covering,cho2010reweighted}. Pairwise constraints have shown to be a popular alternative for cases that require more flexibility such as scenes with moving elements and non-planar perspective change. Introduced by Leordeanu and Herbert \cite{leordeanu2005spectral}, pairwise constraints work by applying a geometric constraint across correspondences on a pairwise basis, attempting to minimize the pairwise error. This approach has later been adopted by Pang et al.\ but made scale invariant and more efficient by using a voting scheme \cite{yuan2012efficient,pang2012scale}. Similarly we can cluster matches together based on a geometric constraint and treat each cluster as a separate case \cite{cho2009feature,wu2011robust}. An interesting application of this principle is demonstrated by Chen et al.\ who iteratively use a Hough transform to cluster matches using angular constraints \cite{chen2013robust}.

@@

\emph{Fast-Match} makes use of an angular assumption to efficiently find new matches in the geometric neighbourhood of already confirmed matches. However this constraint is only applied \emph{locally} to increase the number of matches, making \emph{Fast-Match} robust to outliers. In addition \emph{Fast-Match} does not rely on an initial set of matches and derives much of its speed from the fact that not even an initial set of local image features is required, making it faster than \emph{Ratio-Match} on most occasions.

This paper is structured as follows: Section~\ref{algorithm} introduces the \emph{Fast-Match} algorithm. Section~\ref{experiments} outlines the experimental setup. In Section~\ref{results} we discuss the results before concluding in Section~\ref{conclusion}.

@@

In the worst case, each feature in the \emph{target image} has a true correspondence in the \emph{query image} and $\mathbb{P}(A_{ti} = 0.5)$, in which case we find can find $\mathbb{P}(X_t = n) = \mathbb{P}(A_{ti})^n = 0.5^n$, as well as the expected value $\mathbb{E}(X_t) = \sum_{i=1}^\infty 0.5^ii = 2$, and the variance $Var(X_t) = \sum_{i=1}^\infty 0.5^i(i - 2)^2 = 6$.

@@

For the above case it is theoretically possible that for every one of the $n$ features we end up considering $kn$ seed matches for some constant $k$, i.e.\ a quadratic behaviour. Using Chebyshev's inequality we can provide an upper bound on the probability of this event:
\begin{align}
    \label{bound}
    \mathbb{P}\left(\left| \frac{S_n}{n} - \mathbb{E}(X_t) \right| \ge kn\right) &\le \frac{Var\left(\frac{S_n}{n}\right)}{k^2n^2}\\
&\le \frac{6}{k^2n^3}
\end{align}

@@

\subsection{Evaluation of Fast-Match on 3D Objects}
\label{3dobjects}
%
The 3D Objects dataset by Moreels and Perona \cite{moreels2007evaluation} allows us to experimentally compare matching algorithms over a large range of object and surface types rotated on a turnstile and photographed from every angle in increments of 5 degrees.  15 objects from the dataset are shown in Figure~\ref{fig:3d_objects}.  We use images of 84 different objects under three different lighting conditions at 12 different angle intervals, conducting experiments with a total of 3024 image pairs.

@@

To compute the total number of possible correspondences, we take each feature in a \emph{query image} and count how many of them have a feature in the \emph{target image} which would satisfy the epipolar constraints. When using this dataset, features with no correspondences were not included in the set of features for testing, so as to avoid matching non-moving background and foreground objects.

% TODO: do we still match up to 60?
We evaluate all matching algorithms from our framework on the 3D Objects dataset by matching images at different angular intervals. For each object we pick the \emph{query} image as the image taken at 10 degrees rotation for calibration stability.  We then match this image with the same object turned an additional $\Delta$ degrees, $\Delta \in \{5, 10, \ldots, 60\}$.  For every angle interval we compare images taken under 3 different lighting conditions as provided by the dataset. We include all objects in the database for which photos at 5 degree angle intervals are available except for the ``Rooster'' and ``Sponge'' objects due to image irregularities.

@@

To achieve a good balance between speed and robustness we let the region in which we extract features be a square with a side length of 90 pixels. On empirical tests we found that anything smaller would result in decreased performance while much bigger regions would decrease speed. The 90 pixel window is split in to nine smaller squares. When a seed match falls in any of these squares we match only the features within the smaller square. We let both the regions and the smaller squares overlap each other at all edges with 25 pixels in order to capture feature points lying close to an edge. For the image we have already cached we find all features within a radius of 50 pixels of the seed match.

@@

The high recall rates of \emph{Fast-Match} are partially explained by the fact that we can allow our confidence thresholds to be more lenient when we already know that we are likely to find correct correspondences in the regions that we are matching. However, lenient thresholds can easily impact the matching speed, so in practical situations we usually have to choose between matching fast and precisely, or more slowly with higher recall.

Figure~\ref{fig:match_example} and \ref{fig:timings} demonstrate the speed of \emph{Fast-Match} on image sizes as is typically produced by modern consumer cameras and camera phones. In Figure~\ref{fig:timings} we compare the speed of the two variations of \emph{Fast-Match} to different variations of \emph{Ratio-Match} on two image pairs, one with two images of 10 megapixel and another where the same image pair has been scaled to 3 megapixels. For the larger images we see the clearest difference in performance. The retrieval variant of \emph{Fast-Match} matches the image pairs in around one second while a retrieval variant of \emph{Ratio-Match} with precomputed features spends eight seconds on the same task. This result is roughly equal to the time the general variation of \emph{Fast-Match} spends matching the two images without any pre-computations. For the nearest neighbor search we note that for large images there are substantial speed gains to be had by choosing the right algorithm for finding nearest neighbors, with Muja et al.'s Flann matcher being vastly superior to brute force and KGraph.

>>>>

In this paper we introduce a matching algorithm designed to match features between two images only in image areas that are likely to correspond. This approach is much faster than traditional methods because there is no need to compute descriptors for areas in the image that are not matched. We provide two variations of the algorithm: The \emph{general} variation functions as a traditional matching algorithm and matches two unknown images albeit faster and more robustly than conventional methods. The \emph{retrieval} variation on the other hand assumes that we know one of the images we intend to match beforehand; under this assumption, it can be a magnitude faster than existing matching methods without compromising on accuracy. We will unimaginatively refer to the proposed algorithm as \emph{Fast-Match} in this paper and make clear from the context whether this refers to the \emph{retrieval} or \emph{general} variation.

@@

\emph{Fast-Match} makes use of an angular assumption to efficiently find new matches in the geometric neighbourhood of already confirmed matches. However this constraint is only applied \emph{locally} to increase the number of matches, making \emph{Fast-Match} robust to outliers. In addition \emph{Fast-Match} does not rely on an initial set of matches and derives much of its speed from the fact that not even an initial set of local image features is required.

This paper is structured as follows: Section~\ref{related} discuss related work. Section~\ref{algorithm} introduces the \emph{Fast-Match} algorithm. Section~\ref{experiments} outlines the experimental setup. In Section~\ref{results} we discuss the results before concluding in Section~\ref{conclusion}.

\section{Related work}
\label{related}
%
As noted, the cost of computing SIFT keypoints and descriptors is addressed by other local image features such as SURF \cite{bay2006surf}, BRIEF \cite{calonder2010brief}, or BRISK \cite{leutenegger2011brisk}, just to mention a few. Similarly efforts have been made to improve SIFT itself such as PCA-SIFT \cite{ke2004pca} and GLOH \cite{mikolajczyk2005performance}. Both apply PCA to reduce descriptor lengths and improve distinctiveness, but neither have been shown to consistently outperform SIFT 
\cite{mikolajczyk2005performance,li2008comprehensive}.

@@

A wealth of matching methods have focused instead on increasing the efficiency of local feature matching. \emph{Fast-Match} builds upon the foundation of \emph{Ratio-Match} introduced originally by Deriche et al.~\cite{deriche1994robust} and Baumberg~\cite{baumberg2000reliable} even though Lowe is usually credited for introducing it~\cite{lowe2004sift}. They both propose using the ratio of the similarity of the best to second best match of a given point to evaluate how unique it is. Their finding has later been tested by several independent teams, all concluding that thresholding based on this ratio is generally superior to thresholding based on similarity or returning all nearest neighbours \cite{lowe2004sift,mikolajczyk2005performance,moreels2007evaluation,rabin2009statistical}.

Brown and Lowe \cite{brown2005multi} extend ratio match to deal with a set of images by using not the ratio of the best and second best match, but the average ratio of the best and the average of second best matches across a set of images.  Rabin et al.\ \cite{rabin2009statistical} try to enhance descriptor matching by looking at the statistical distribution of local features in the matched images, and only return a match when such a correspondence would not occur by mere chance. Finally, Arnfred et al.\ generalize \emph{Ratio-Match} to make use of both of the matched images to provide a more accurate evaluation of the uniqueness of a match \cite{arnfred2013mirror}.

@@

For sparse local image features many solutions have combined \emph{Ratio-Match} with various geometric constraints to improve matching. These constraints are based on assumptions regarding the transformation between the \emph{query} and \emph{target images}. A commonly used example is \emph{RANSAC} applied to feature matching where matches are chosen from a pool of candidates according to how well they approximate a global epipolar geometry \cite{fischler1981ransac,torr2000mlesac,szeliski2010}. Similar global angular and distance constraints can be used to filter a set of matches as shown in \cite{kim2008efficient,schmid1997local}. Finally the problem of feature matching can be modelled as a graph matching problem where each feature is a vertex, and edge values correspond to a geometric relation between two features as demonstrated in \cite{torresani2008feature,yarkony2010covering,cho2010reweighted}. For cases that require more flexibility such as scenes with moving elements and non-planar perspective change Pairwise constraints provide an alternative to the more rigid global constraints. Introduced by Leordeanu and Herbert \cite{leordeanu2005spectral}, pairwise constraints work by applying a geometric constraint across matches on a pairwise basis, attempting to minimize the pairwise error. This approach has later been adopted by Pang et al.\ but made scale invariant and more efficient by using a voting scheme \cite{yuan2012efficient,pang2012scale}. Similarly we can cluster matches together based on a geometric constraint and treat each cluster as a separate case \cite{cho2009feature,wu2011robust}. An interesting application of this principle is demonstrated by Chen et al.\ who iteratively use a Hough transform to cluster matches using angular constraints \cite{chen2013robust}.

@@

@@

In the worst case, each feature in the \emph{target image} has a true correspondence in the \emph{query image} and $\mathbb{P}(A_{ti} = 0.5)$, in which case we find can find $\mathbb{P}(X_t = n) = \mathbb{P}(A_{ti})^n = 0.5^n$, as well as the expected value $\mathbb{E}(X_t) = \sum_{i=1}^\infty i0.5^i = 2$, and the variance $Var(X_t) = \sum_{i=1}^\infty 0.5^i(i - 2)^2 = 6$.

@@

For the above case it is theoretically possible that for every one of the $n$ features we end up considering $kn$ seed matches for some constant $k$, i.e.\ a quadratic behaviour. Using Chebyshev's inequality we can provide an upper bound on the probability of this event:
\begin{align}
    \label{bound}
    \mathbb{P}\left(\left| \frac{S_n}{n} - \mathbb{E}(X_t) \right| \ge kn\right) &\le \frac{Var\left(\frac{S_n}{n}\right)}{k^2n^2} = \frac{6}{k^2n^3}
\end{align}

@@

\subsection{Evaluation of Fast-Match on 3D Objects}
\label{3dobjects}
%
The 3D Objects dataset by Moreels and Perona \cite{moreels2007evaluation} allows us to experimentally compare matching algorithms over a large range of object and surface types rotated on a turnstile and photographed from every angle in increments of 5 degrees. 15 objects from the dataset are shown in Figure~\ref{fig:3d_objects}.  We use images of 84 different objects photographed under three different lighting conditions at 12 different angle intervals, conducting experiments with a total of 3024 image pairs. All photos a taken with a consumer camera in 3.1 mega pixel resolution.

@@

To compute the total number of possible true correspondences, we take each feature in a \emph{query image} and count how many of them have a feature in the \emph{target image} which would satisfy the epipolar constraints. When using this dataset, features with no correspondences were not included in the set of features for testing, so as to avoid matching non-moving background and foreground objects.

We evaluate all matching algorithms from our framework on the 3D Objects dataset by matching images at different angular intervals. For each object we pick the \emph{query} image as the image taken at 10 degrees rotation for calibration stability.  We then match this image with the same object turned an additional $\Delta$ degrees, $\Delta \in \{5, 10, \ldots, 60\}$.  For every angle interval we compare images taken under 3 different lighting conditions as provided by the dataset. We include all objects in the database for which photos at 5 degree angle intervals are available except for the ``Rooster'' and ``Sponge'' objects due to image irregularities.

@@

To achieve a good balance between speed and robustness we let the region in which we extract features be a square with a side length of 90 pixels. On empirical tests we found that anything smaller would result in decreased performance while much bigger regions would decrease speed. The 90 pixel window is split in to nine smaller squares (see also Figure~\ref{fig:exploration}). When a seed match falls in any of these squares we match only the features within the smaller square. We let both the regions and the smaller squares overlap each other at all edges with 25 pixels in order to capture feature points lying close to an edge. For the image we have already cached we find all features within a radius of 50 pixels of the seed match.

@@

The high recall rates of \emph{Fast-Match} are partially explained by the fact that we can allow our confidence thresholds to be more lenient when we already know that we are likely to find true correspondences in the regions that we are matching. However, lenient thresholds can easily impact the matching speed, so in practical situations we usually have to choose between matching fast and precisely, or more slowly with higher recall.

Figure~\ref{fig:timings} demonstrate the speed of \emph{Fast-Match} on image sizes as is typically produced by modern consumer cameras and camera phones. In Figure~\ref{fig:timings} we compare the speed of the two variations of \emph{Fast-Match} to different variations of \emph{Ratio-Match} on two image pairs, one with two images of 10 megapixel and another where the same image pair has been scaled to 3 megapixels. For the larger images we see the clearest difference in performance. The retrieval variant of \emph{Fast-Match} matches the image pairs in around one second while a retrieval variant of \emph{Ratio-Match} with precomputed features spends eight seconds on the same task. This result is roughly equal to the time the general variation of \emph{Fast-Match} spends matching the two images without any pre-computations. For the nearest neighbor search we note that for large images there are substantial speed gains to be had by choosing the right algorithm for finding nearest neighbors, with Muja et al.'s Flann matcher being vastly superior to brute force and KGraph.
